MetaTracts - A Method for Robust Extraction and Visualization of Carbon Fiber Bundles in Fiber Reinforced Composites

### 2014-09-24 chapters/MetaTracts.tex
reviewed until user eval

\begin{figure}[tb]
\centering
	\includegraphics[width=0.45\textwidth]{images_pvis/figure5.png}
	%\includegraphics[width=0.45\textwidth,  trim = 0mm 100mm 0mm 50mm, clip]{imagesMT2014/image_length}
	\caption{(A) MetaTracts colored according to mean orientations. (B) length distribution of individual MetaTracts for a particular bundle.(unit for length is grid cube edge length:  $2\mu m$). A single fiber bundle is made of MetaTracts of varying length}
	\label{fig:length_distribution}
\end{figure}

@@

The output of the above step is a set of MetaTracts. The next step is to cluster the MetaTracts into bundles. 
Carbon fibers have both orientation and geometric proximity information, both of which can be used for clustering. Experimentally we found that orientation measure was very reliable while measures of geometric proximity had problems with fibers which partially overlap (Fig.~\ref{fig:length_distribution}B).  
Instead of creating a heuristic which artificially combines the orientation and geometric proximity measures, we first cluster based on orientation and then further subdivide each orientation cluster using geometric proximity.
We found that different clustering techniques performed preferably for different measures.  The two step clustering also made the problem more tractable and  the parameters intuitive for the end users. For the orientation clustering, we use dimension reduction followed by K-means clustering. We then use hierarchical clustering to further subdivide each cluster based on the geometric proximity measures.

@@

\subsection {Orientation based clustering}
%We use the inherent features expressed as weaving pattern of our data to pre-cluster the metaTracts into the number of classes based on the major directions of the pattern.
%It is important to note that this step broadly divides the MetaTracts into classes based on the major orientations and not individual fiber bundles. 
This step divides the individual MetaTracts into classes based on their major orientations. In order to cluster MetaTracts going in same directions, we use a spectral embedding technique called Laplacian eigenmaps which was originally introduced by Belkin and Niyogi \cite{Belkin01} and later used in DTI fiber coloring~\cite{Brun2003} and fiber clustering~\cite{Brun2004}. The key notion is to find a suitable similarity measure to define the weights of an undirected weighted graph. The nodes being the data points and the edges being the weights. Then, an eigenvalue problem is solved, which maps the manifold embedded in the graph into a lower dimensional space while preserving the graph structure.
This lower dimensional space can then be clustered more efficiently than the higher dimensional space. 
Let G be the graph, we compute the eigenvalues and eigenvectors for the generalized eigenvector problem:$L\textbf{f}=\lambda D\textbf{f}$,
%\begin{equation}\label{equn:eigenMaps}
%L\textbf{f}=\lambda D\textbf{f}
%\end{equation}
where D is the diagonal weight matrix and L is the Laplacian matrix. The eigenvector \textbf{${f}_{0}$} corresponding to the eigenvalue 0 is left out and the next $m$, {\textbf{${f}_{1}$} through \textbf{${f}_{m}$}} eigenvectors are used to embed in an m-dimensional space. K-means clustering is then used to cluster in this lower dimensional space.

Brun et al.~\cite{Brun2003} based their measure on the simple assumption that two traces with similar endpoints should be considered similar. Brun et al. ~\cite{Brun2004} based their measure on a 9-D fiber descriptors. In our case, each MetaTract is a data point. We introduce a simple \textit{orientation based similarity measure}. There is no spatial information involved, just a partition of the MetaTracts based on orientation which is already inherent in the data.
Following this, given a pair of  MetaTracts, we define the edge weights between two tracts as the cosine of the maximum angle between the local orientations ($N_P$) of all pairs of start points ($C_P$) between the two MetaTracts. The edge weights give a distance matrix representing the distance between each pair of points. Using the Belkin and Niyogi algorithm we ``embed" these points in a low dimensional space where the Euclidean distance between points approximates the distance between points given by the distance matrix. 
%
%August 13 2014
%For all our tests the fibers were then embedded in a $\eta$ low dimensional space. The number of dimensions in the high dimensional space is equal to the number of datapoints or the number of MetaTracts in our case.
%For all our test cases we set $\eta$ to be five (see sec~\ref{sec:param_choices} for parameter choices for $\eta$.).
%  
We used conventional K-means for clustering this lower dimensional space. Here $K$ is supplied by domain knowledge of the number of major fiber bundle directions of the woven structure.
For our test case there are two major directions of the fiber bundles and $K$ was set to 2. Due to the advantages offered by the dimensionality reduction, even if there are curved fibers bundles the user has to provide just the major fiber bundle directions of the weaving pattern. Figure~\ref{fig:orientation_clustering}B shows the result of the K-means clustering with the data points projected to the top three eigenvectors as the major axes. As is expected there is a clear distinction based on fiber orientation. Figure~\ref{fig:orientation_clustering}(A,C) shows the MetaTracts colored according to clustering results.

@@

\subsection{Distance based clustering}
 \label{subsec:dist_clustering}
Orientation clustering uses only orientation information. To subdivide the oriented clusters into fiber bundles, we need to include further information about the geometric proximity between MetaTracts. We use the directed Hausdorff distance for distance based clustering.
Each MetaTract is represented as a set of points ($C_P$). Formally, the directed Hausdorff distance from point set P to point set Q is defined as 
$H_{dir}(P,Q) = max_{p \in P} min_{q \in Q} d(p,q)$ .
The Hausdorff distance is defined as $H(P,Q) = max(H_{dir}(P,Q),H_{dir}(Q,P))$.
The Hausdorff distance is a metric so $H(P,Q) \le H(P,Q') + H(Q',Q)$ but the directed Hausdorff is not.
Unfortunately, the Hausdorff distance does not work well for our application since a fiber bundle may have many MetaTract which only partially covers the bundle (Figure~\ref{fig:length_distribution}(B)). If a MetaTract $P$ covers only part of the fiber bundle covered by $Q$, then $H_{dir}(P,Q)$ will be very small while $H_{dir}(Q,P)$ will be large.
Thus, $H(P,Q)$ will be large, even though $P$ and $Q$ are in the same fiber bundle.
Instead of using the Hausdorff distance, $\max(H_{dir}(P,Q),H_{dir}(Q,P)$, we use $\min(H_{dir}(P,Q),H_{dir}(Q,P))$. If $P$ covers only part of the fiber bundle covered by $Q$, then $\min(H_{dir}(P,Q),H_{dir}(Q,P))$ is very small.
Note that if $P$ and $Q$ overlap but do not cover the same parts of the fiber bundle, then $H_{dir}(P,Q)$ and $H_{dir}(Q,P)$ and $\min(H_{dir}(P,Q),H_{dir}(Q,P))$ will be large.

The directed Hausdorff distance is very sensitive to outliers in the data.
However, because MetaTracts after orientation clustering are constructed using cylinders with similar orientations, they are not plagued by outliers.
To cluster based on MetaTract proximity, we used single linkage hierarchical clustering.
Hierarchical clustering has a single parameter $h$, the desired number of clusters.
Clusters are merged until there are only $h$ clusters left.
Hierarchical clustering is intuitive since it is easy to trace how clusters are formed and merged.
Single linkage clustering finds pairs of objects $p \in P$ and $q \in Q$ where $P \neq Q$ which are closer than other such pairs, and merges the containing clusters $P$ and $Q$.
We found that single linkage hierarchical clustering had two major drawbacks.
First, the clustering would produce some small clusters of just a few MetaTracts.
These MetaTracts were anomalies caused by overlapping fibers and did not represent true fiber bundles.
Second, if two fiber bundles were parallel for some of their length and then separated, they would sometimes be clustered into the same fiber bundle.
A short MetaTract which was parallel to both and did not extend into the separation region could form a link between the two fiber bundles, causing them to be clustered into a single bundle.
To address the problem of small clusters, we applied hierarchical clustering and then identified small clusters with few MetaTracts.
We removed the MetaTracts that were in those clusters from the data set and reapplied hierarchical clustering.
To address the problem of short MetaTracts joining different fiber bundles, we applied hierarchical clustering and then removed the
shortest fibers (fibers less than 0.6 times the median length) in each bundle. We then reapplied hierarchical clustering.
We repeated both until a steady state of clusters is reached and no new small fibers can be removed. The results of clustering the orientation clusters Fig.~\ref{fig:orientation_clustering}(A,C) are shown in Fig.~\ref{fig:orientation_clustering}(D,E) respectively. Note how the thin and curved purple cluster Fig.~\ref{fig:orientation_clustering}(E) bundle ``7'' is extracted well.

In comparison Fig.~\ref{fig:comparison} shows the MetaTracts in Fig.~\ref{fig:length_distribution}(A) being hierarchically clustered directly using proximity alone into 10,15 and 20 clusters (ground truth being 11). As discussed above, single linkage hierarchical clustering in the presence of overlapping fibers tends to create large clusters and small, low cardinality outliers. In Fig.~\ref{fig:comparison} when ``h" is 10 two large multi-bundle incorrect cluster is generated and the rest are outliers. As ``h" increases, some appropriate bundles start to form. Even at ``h" equals 20, the top bundles all clusters together into one large incorrectly identified bundle.
\begin{figure}
\centering
		\includegraphics[width=0.4\textwidth]{images_pvis/figure7}
		%\includegraphics[width=0.5\textwidth, trim = 0mm 70mm 0mm 20mm, clip]{imagesMT2014/image7b.pdf}
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
	\caption{(A) Volume rendering of data set Figure~\ref{fig:data-char}A.(B) shows a single slice of the data set. (C) shows two of the extracted meshes together. (D) shows the meshes separately.}
	\label{fig:crop-16-decomp}
\end{figure}

@@

\section {Voxelization and surface extraction}
\label{sec:vis}
Apart from direct visualization of the MetaTracts, we show two simple yet highly important extensions for our domain specialist. The first one is to voxelize the original volume according to the clusters each voxel is associated with.
The second is to extract the corresponding surfaces from the voxel data by binarizing the volume per cluster and extracting the isosurface of the largest connected component from the binary volume. Both methods can possibly be used for further analysis of the data, e.g. in simulations.
To voxelize the space,  we compute a neighborhood around each voxel. We then enumerate the number of voxels of each class (cluster) in this neighborhood. The voxel is then assigned to the class with the maximum number of elements in the neighborhood. Figure~\ref{fig:crop-16-decomp}A and Figure~\ref{fig:prepreg}D shows the result of voxelization. Figure~\ref{fig:crop-16-decomp}(C, D) show examples of extracted meshes.

@@

 %glass
We tested our method on multiple datasets with varying characteristics. 
Figure~\ref{fig:orientation_clustering} shows the result of data set~\ref{fig:data-char} decomposed into 2 orientation and each orientation decomposed into ``h" equals 10 clusters (the ground truth is 6 and 5 respectively). The hierarchical clustering is robust to over segmentation.
Figure~\ref{fig:len_dist_crop16} shows the median, minimum and maximum lengths per cluster. We generate the correct clusters and outlier bundles with very few elements which can be discarded. As noted in section~\ref{subsec:dist_clustering} the tracts do not run the entire length of the dataset, the median length of MetaTracts in cluster 6 (Fig.~\ref{fig:length_distribution}) is 200 while the maximum MetaTract length in the bundle is about 500 (measured in units of grid cube length: $2\mu m$). Fig.~\ref{fig:length_distribution}(B) shows the length per MetaTract distribution for a particular fiber bundle. 
Fig.~\ref{fig:prepreg} shows the results on our dataset2. This dataset is dense with \textit{flat and thin} bundles. The dimensions are 300x350x300 and the datatype is uint16 (compared to dataset1 which was only uint8). The bundles are indistinguishable in the original data set Fig.~\ref{fig:prepreg}(A). A green dotted line shows one of the fiber bundle boundary. Fig.~\ref{fig:prepreg}(B) shows the fiber bundles for the two orientation. Fig.~\ref{fig:prepreg}(C) shows the combined results.Fig.~\ref{fig:prepreg}(D) shows the volume after voxelization.
The fiber bundle extraction using MetaTracts was implemented in C++ using ITK. All clustering was done in R. The preprocessing consists of MetaTract generation the distance computation and the orientation clustering. The hierarchical clustering separately, on the results of the orientation clusters can be done quickly in a few minutes on a Intel Xeon E5-2667 workstation according to user input `h'(section~\ref{sec:param_choices}). We store the normalized direction at each fiber point, thus both distance measures can be extracted simultaneously without extra computation.

@@

\section{Parameter choices}
\label{sec:param_choices}
\begin{figure}
\centering
	\includegraphics[width=0.45\textwidth,  trim = 0mm 260mm 0mm 15mm, clip]{images_pvis/figure9}
	%\includegraphics[width=0.5\textwidth,  trim = 10mm 40mm 0mm 15mm, clip]{imagesMT2014/image_length_histogram.pdf}
	\caption{Number of tracts and median, minimum and maximum length of individual MetaTracts in the orientation cluster Fig.~\ref{fig:orientation_clustering}A, clustered into 10 clusters. The unit for length is the grid cube length. Figure~\ref{fig:length_distribution}(B) shows the length of individual MetaTracts in cluster 6. Clusters labeled 2,3,4,5 and 7 all have cardinality less than 10. }
	\label{fig:len_dist_crop16}
\end{figure}
The critical parameters are $K$ for the K-means in orientation based clustering and $h$ for the hierarchical clustering. K denotes the number of major directions, which is either known a priori or can be easily estimated by looking at the weaving pattern. 
Our framework is robust to the choice of $h$ (Section~\ref{sec:results}). Large $h$ keeps the major bundles intact. For example, in  Fig.~\ref{fig:len_dist_crop16}, where $h$ was set to ten while the ground truth was five, we observe that the major clusters remain well segmented and small clusters can be removed if they have too few elements (clusters 2,3,4,5 and 7 in Fig.~\ref{fig:len_dist_crop16} all have less than 10 elements). This is an appealing trait of the proximity based clustering, thus providing good results even when exact $h$ might be unknown.
The following parameters were fixed for all the tests. We set the reliable Hessian threshold $R_{h}$ to be 0.3 for all our tests. A $R_{h}$ of 0.0 would mean all points have reliable local orientation which would cause spurious MetaTracts detection. On the other hand a very high $ R_{h}$ would lead to be a decline in number of MetaTracts produced. The $\alpha$ and $\beta$ in $R_{h}$ are as explained in Frangi et al. \cite{Frangi1998} and set to 0.5. The length and the width parameters for the cylinders for the MetaTracts decide how coarse the fiber cylinders are. Larger cylinders will handle noisy local orientation better as it inspects a higher number of candidate points to extend the fiber. We used 10.00 and 2.00 for length and breadth(measured in grid cube length), respectively for all tests. $\alpha$ and $\beta$ in equation~\ref{eqn:algo_1} decide how quickly the value of the factor decays, we have used [7-10] and half the length of cylinder (5.0). Our number of fiber bundle directions are limited, thus even for small $m$ (cardinality of lower dimension in orientation clustering), the distinction between the orientation clusters is preserved quite well. We compared  $m=3-7$ experimentally without any dramatic change in results. 
\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth, trim = 0mm 80mm 0mm 170mm, clip]{images_pvis/figure10}
	\caption{Direct hierarchical clustering of MetaTracts Fig.~\ref{fig:length_distribution}(A) by proximity into 10,15 and 20 clusters.}
	\label{fig:comparison}
\end{figure} 
%\begin{figure}
%\centering
% \subfloat[$\eta=7$]{\includegraphics[width=0.3\linewidth]{imagesMetaTracts/Rplot-ndim3k2}}
%  \subfloat[$\eta=3$]{\includegraphics[width=0.3\linewidth]{imagesMetaTracts/Rplot-ndim7k2}}
%  \caption{Embedding from $\mathbb{R}^m\to \mathbb{R}^\eta$, here m is the number of MetaTracts, 580 for this example. Followed by clustering into K=2 clusters.}
%  \label{fig:eta}
%\end{figure}
%\begin{figure}[htp]
%\centering
% \includegraphics[width=\linewidth]{imagesMetaTracts/kul-results.eps}
%  \caption{Results from orientation based dimensionality reduction and clustering. A is the original data set, B shows the results of fiber tracking. C shows the results of orientation based clustering and D shows the results of the proximity based clustering. }\label{fig:kull-full}
%\end{figure}

>>>>

\begin{figure}[tb]
\centering
	\includegraphics[width=0.45\textwidth]{images_pvis/figure5.png}
	%\includegraphics[width=0.45\textwidth,  trim = 0mm 100mm 0mm 50mm, clip]{imagesMT2014/image_length}
	\caption{(A) MetaTracts color-coded according to their mean orientations. (B) length distribution of individual MetaTracts for a particular bundle (unit for length is grid cube edge length:  $2\mu m$). A single fiber bundle is made of MetaTracts of varying length}
	\label{fig:length_distribution}
\end{figure}

@@

The output of the above step is a set of MetaTracts. The next step is to cluster the MetaTracts into bundles. 
Carbon fibers have both orientation and geometric proximity information, both of which can be used for clustering. Experimentally we found that orientation measure was very reliable while measures of geometric proximity had problems with fibers which partially overlap (Figure~\ref{fig:length_distribution}b).  
Instead of creating a heuristic which artificially combines the orientation and geometric proximity measures, we first cluster based on orientation and then further subdivide each orientation cluster using geometric proximity.
We found that different clustering techniques performed preferably for different measures.  The two step clustering also made the problem more tractable and  the parameters intuitive for the end users. For the orientation clustering, we use dimension reduction followed by K-means clustering. We then use hierarchical clustering to further subdivide each cluster based on the geometric proximity measures.

@@

\subsection {Orientation based clustering}
%We use the inherent features expressed as weaving pattern of our data to pre-cluster the metaTracts into the number of classes based on the major directions of the pattern.
%It is important to note that this step broadly divides the MetaTracts into classes based on the major orientations and not individual fiber bundles. 
This step divides the individual MetaTracts into classes based on their major orientations. In order to cluster MetaTracts going in same directions, we use a spectral embedding technique called Laplacian eigenmaps which was originally introduced by Belkin and Niyogi \cite{Belkin01} and later used in DTI fiber coloring~\cite{Brun2003} and fiber clustering~\cite{Brun2004}. The key notion is to find a suitable similarity measure to define the weights of an undirected weighted graph. The nodes being the data points and the edges being the weights. Then, an eigenvalue problem is solved, which maps the manifold embedded in the graph into a lower dimensional space while preserving the graph structure.
This lower dimensional space can then be clustered more efficiently than the higher dimensional space. 
Let $G$ be the graph, we compute the eigenvalues and eigenvectors for the generalized eigenvector problem:$L\textit{\textbf{f}}=\lambda D\textit{\textbf{f}}$,
%\begin{equation}\label{equn:eigenMaps}
%L\textbf{f}=\lambda D\textbf{f}
%\end{equation}
where $D$ is the diagonal weight matrix and $L$ is the Laplacian matrix. The eigenvector \textbf{${f}_{0}$} corresponding to the eigenvalue 0 is left out and the next $m$, {\textbf{${f}_{1}$} through \textbf{${f}_{m}$}} eigenvectors are used to embed in an $m$-dimensional space. K-means clustering is then used to cluster in this lower dimensional space.

Brun et al.~\cite{Brun2003} based their measure on the simple assumption that two traces with similar endpoints should be considered similar. Brun et al. ~\cite{Brun2004} based their measure on a 9-D fiber descriptors. In our case, each MetaTract is a data point. We introduce a simple \textit{orientation based similarity measure}. There is no spatial information involved, just a partition of the MetaTracts based on orientation which is already inherent in the data.
Following this, given a pair of  MetaTracts, we define the edge weights between two tracts as the cosine of the maximum angle between the local orientations ($N_P$) of all pairs of start points ($C_P$) between the two MetaTracts. The edge weights give a distance matrix representing the distance between each pair of points. Using the Belkin and Niyogi algorithm we ``embed" these points in a low dimensional space where the Euclidean distance between points approximates the distance between points given by the distance matrix. 
%
%August 13 2014
%For all our tests the fibers were then embedded in a $\eta$ low dimensional space. The number of dimensions in the high dimensional space is equal to the number of datapoints or the number of MetaTracts in our case.
%For all our test cases we set $\eta$ to be five (see sec~\ref{sec:param_choices} for parameter choices for $\eta$.).
%  
We used conventional K-means for clustering this lower dimensional space. Here $K$ is supplied by domain knowledge of the number of major fiber bundle directions of the woven structure.
For our test case there are two major directions of the fiber bundles and $K$ was set to 2. Due to the advantages offered by the dimensionality reduction, even if there are curved fibers bundles the user has to provide just the major fiber bundle directions of the weaving pattern. Figure~\ref{fig:orientation_clustering}b shows the result of the K-means clustering with the data points projected to the top three eigenvectors as the major axes. As is expected there is a clear distinction based on fiber orientation. Figure~\ref{fig:orientation_clustering}a,c shows the MetaTracts colored according to clustering results.

@@

\subsection{Distance based clustering}
 \label{subsec:dist_clustering}
Orientation clustering uses only orientation information. To subdivide the oriented clusters into fiber bundles, we need to include further information about the geometric proximity between MetaTracts. We use the directed Hausdorff distance for distance based clustering.
Each MetaTract is represented as a set of points ($C_P$). Formally, the directed Hausdorff distance from point set $P$ to point set $Q$ is defined as 
$H_{dir}(P,Q) = max_{p \in P} min_{q \in Q} d(p,q)$ .
The Hausdorff distance is defined as $H(P,Q) = max(H_{dir}(P,Q),H_{dir}(Q,P))$.
The Hausdorff distance is a metric so $H(P,Q) \le H(P,Q') + H(Q',Q)$ but the directed Hausdorff is not.
Unfortunately, the Hausdorff distance does not work well for our application since a fiber bundle may have many MetaTract which only partially covers the bundle (Figure~\ref{fig:length_distribution}b). If a MetaTract $P$ covers only part of the fiber bundle covered by $Q$, then $H_{dir}(P,Q)$ will be very small while $H_{dir}(Q,P)$ will be large.
Thus, $H(P,Q)$ will be large, even though $P$ and $Q$ are in the same fiber bundle.
Instead of using the Hausdorff distance, $\max(H_{dir}(P,Q),H_{dir}(Q,P)$, we use $\min(H_{dir}(P,Q),H_{dir}(Q,P))$. If $P$ covers only part of the fiber bundle covered by $Q$, then $\min(H_{dir}(P,Q),H_{dir}(Q,P))$ is very small.
Note that if $P$ and $Q$ overlap but do not cover the same parts of the fiber bundle, then $H_{dir}(P,Q)$ and $H_{dir}(Q,P)$ and $\min(H_{dir}(P,Q),H_{dir}(Q,P))$ will be large.

The directed Hausdorff distance is very sensitive to outliers in the data.
However, because MetaTracts after orientation clustering are constructed using cylinders with similar orientations, they are not plagued by outliers.
To cluster based on MetaTract proximity, we used single linkage hierarchical clustering.
Hierarchical clustering has a single parameter $h$, the desired number of clusters.
Clusters are merged until there are only $h$ clusters left.
Hierarchical clustering is intuitive since it is easy to trace how clusters are formed and merged.
Single linkage clustering finds pairs of objects $p \in P$ and $q \in Q$ where $P \neq Q$ which are closer than other such pairs, and merges the containing clusters $P$ and $Q$.
We found that single linkage hierarchical clustering had two major drawbacks.
First, the clustering would produce some small clusters of just a few MetaTracts.
These MetaTracts were anomalies caused by overlapping fibers and did not represent true fiber bundles.
Second, if two fiber bundles were parallel for some of their length and then separated, they would sometimes be clustered into the same fiber bundle.
A short MetaTract which was parallel to both and did not extend into the separation region could form a link between the two fiber bundles, causing them to be clustered into a single bundle.
To address the problem of small clusters, we applied hierarchical clustering and then identified small clusters with few MetaTracts.
We removed the MetaTracts that were in those clusters from the data set and reapplied hierarchical clustering.
To address the problem of short MetaTracts joining different fiber bundles, we applied hierarchical clustering and then removed the
shortest fibers (fibers less than 0.6 times the median length) in each bundle. We then reapplied hierarchical clustering.
We repeated both until a steady state of clusters is reached and no new small fibers can be removed. The results of clustering the orientation clusters (see Figure~\ref{fig:orientation_clustering}a,c) are shown in Figure~\ref{fig:orientation_clustering}d,e respectively. Note how the thin and curved purple cluster bundle 7 is extracted well (Figure~\ref{fig:orientation_clustering}e).

In comparison Figure~\ref{fig:comparison} shows the MetaTracts in Figure~\ref{fig:length_distribution}a being hierarchically clustered directly using proximity alone into 10, 15, and 20 clusters (ground truth is 11). As discussed above, single linkage hierarchical clustering in the presence of overlapping fibers tends to create large clusters and small, low-cardinality outliers. When number of clusters ($h$) is 10 two large multi-bundle incorrect clusters are generated and the rest are outliers. As $h$ increases, some appropriate bundles start to form. Even at $h=20$, the top bundles all clusters together into one large incorrectly identified bundle.
\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth, trim = 0mm 80mm 0mm 170mm, clip]{images_pvis/figure10}
	\caption{MetaTracts data (left) and its straight-forward hierarchical clustering for various numbers of clusters.}
	\label{fig:comparison}
\end{figure}

@@

\section {Voxelization and surface extraction}
\label{sec:vis}

Apart from direct visualization of the MetaTracts, we show two additional extensions which were requested by our domain specialists as highly important. The first one is voxelizing the original volume according to the clusters each voxel is associated with.
The second is to extract the corresponding surfaces from the voxel data by binarizing the volume per cluster and extracting the isosurface of the largest connected component from the binary volume. Both methods can possibly be used for further analysis of the data, e.g. in simulations.
To voxelize the space,  we compute a neighborhood around each voxel. We then enumerate the number of voxels of each class (cluster) in this neighborhood. The voxel is then assigned to the class with the maximum number of elements in the neighborhood. Figures ~\ref{fig:crop-16-decomp}a and~\ref{fig:prepreg}d show the result of voxelization. Figure~\ref{fig:crop-16-decomp}c,d shows examples of extracted meshes.

\begin{figure}
\centering
		\includegraphics[width=0.4\textwidth]{images_pvis/figure7}
		%\includegraphics[width=0.5\textwidth, trim = 0mm 70mm 0mm 20mm, clip]{imagesMT2014/image7b.pdf}
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
	\caption{(A) Volume rendering of data set Figure~\ref{fig:data-char}A.(B) shows a single slice of the data set. (C) shows two of the extracted meshes together. (D) shows the meshes separately.}
	\label{fig:crop-16-decomp}
\end{figure}

@@

 %glass
We tested our method on multiple datasets with varying characteristics. 
Figure~\ref{fig:orientation_clustering} shows the result of dataset~\ref{fig:data-char} decomposed into 2 orientation clusters with each orientation orientation cluster decomposed into $h=10$ clusters (the ground truth is 6 and 5 respectively). The hierarchical clustering is robust to over segmentation.
Figure~\ref{fig:len_dist_crop16} shows the median, minimum and maximum lengths per cluster. We generate the correct clusters and outlier bundles with very few elements which can be discarded. As noted in section~\ref{subsec:dist_clustering} the tracts do not run the entire length of the dataset, the median length of MetaTracts in cluster 6 (Figure~\ref{fig:length_distribution}) is 200 while the maximum MetaTract length in the bundle is about 500 (measured in units of grid cube length: 2 $\mu m$). Figure~\ref{fig:length_distribution}b shows the length per MetaTract distribution for a particular fiber bundle. 
Figure~\ref{fig:prepreg} shows the results on our dataset 2. This dataset is dense with \textit{flat and thin} bundles. The dimensions are 300$\times$350$\times$300 and the datatype is uint16 (dataset 1 is uint8). The bundles are indistinguishable in the original data set Figure~\ref{fig:prepreg}a. A green dotted line shows one of the fiber bundle boundary. Figure~\ref{fig:prepreg}b shows the fiber bundles for the two orientations, Figure~\ref{fig:prepreg}c shows the combined results, and Figure~\ref{fig:prepreg}d shows the volume after the voxelization.
The fiber bundle extraction using MetaTracts was implemented in C++ using ITK. All clustering was done in R. The preprocessing consists of MetaTract generation the distance computation and the orientation clustering. The hierarchical clustering separately, on the results of the orientation clusters can be done withing an order of a few minutes on Intel Xeon E5-2667 workstation according to user input of $h$ (section~\ref{sec:param_choices}). We store the normalized direction at each fiber point, thus both distance measures can be extracted simultaneously without any extra computations.

@@

\section{Parameter choices}
\label{sec:param_choices}
\begin{figure}
\centering
	\includegraphics[width=0.45\textwidth,  trim = 0mm 260mm 0mm 15mm, clip]{images_pvis/figure9}
	%\includegraphics[width=0.5\textwidth,  trim = 10mm 40mm 0mm 15mm, clip]{imagesMT2014/image_length_histogram.pdf}
	\caption{Number of tracts and median, minimum and maximum length of individual MetaTracts in the orientation cluster Fig.~\ref{fig:orientation_clustering}A, clustered into 10 clusters. The unit for length is the grid cube length. Figure~\ref{fig:length_distribution}(B) shows the length of individual MetaTracts in cluster 6. Clusters labeled 2,3,4,5 and 7 all have cardinality less than 10. }
	\label{fig:len_dist_crop16}
\end{figure}
The critical parameters are $K$ for the K-means in orientation based clustering and $h$ for the hierarchical clustering. K denotes the number of major directions, which is either known a priori or can be easily estimated by looking at the weaving pattern. 
Our framework is robust to the choice of $h$ (Section~\ref{sec:results}). Large $h$ keeps the major bundles intact. For example, in  Fig.~\ref{fig:len_dist_crop16}, where $h$ was set to ten while the ground truth was five, we observe that the major clusters remain well segmented and small clusters can be removed if they have too few elements (clusters 2,3,4,5 and 7 in Fig.~\ref{fig:len_dist_crop16} all have less than 10 elements). This is an appealing trait of the proximity based clustering, thus providing good results even when exact $h$ might be unknown.
The following parameters were fixed for all the tests. We set the reliable Hessian threshold $R_{h}$ to be 0.3 for all our tests. A $R_{h}$ of 0.0 would mean all points have reliable local orientation which would cause spurious MetaTracts detection. On the other hand a very high $ R_{h}$ would lead to be a decline in number of MetaTracts produced. The $\alpha$ and $\beta$ in $R_{h}$ are as explained in Frangi et al. \cite{Frangi1998} and set to 0.5. The length and the width parameters for the cylinders for the MetaTracts decide how coarse the fiber cylinders are. Larger cylinders will handle noisy local orientation better as it inspects a higher number of candidate points to extend the fiber. We used 10.00 and 2.00 for length and breadth(measured in grid cube length), respectively for all tests. $\alpha$ and $\beta$ in equation~\ref{eqn:algo_1} decide how quickly the value of the factor decays, we have used [7-10] and half the length of cylinder (5.0). Our number of fiber bundle directions are limited, thus even for small $m$ (cardinality of lower dimension in orientation clustering), the distinction between the orientation clusters is preserved quite well. We compared  $m=3-7$ experimentally without any dramatic change in results.

%\begin{figure}
%\centering
% \subfloat[$\eta=7$]{\includegraphics[width=0.3\linewidth]{imagesMetaTracts/Rplot-ndim3k2}}
%  \subfloat[$\eta=3$]{\includegraphics[width=0.3\linewidth]{imagesMetaTracts/Rplot-ndim7k2}}
%  \caption{Embedding from $\mathbb{R}^m\to \mathbb{R}^\eta$, here m is the number of MetaTracts, 580 for this example. Followed by clustering into K=2 clusters.}
%  \label{fig:eta}
%\end{figure}
%\begin{figure}[htp]
%\centering
% \includegraphics[width=\linewidth]{imagesMetaTracts/kul-results.eps}
%  \caption{Results from orientation based dimensionality reduction and clustering. A is the original data set, B shows the results of fiber tracking. C shows the results of orientation based clustering and D shows the results of the proximity based clustering. }\label{fig:kull-full}
%\end{figure}
